Rationality, Evolution \& the Environment

### 2015-05-27 paper/paper.tex
section 2 by Paolo

\usepackage[]{helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
\bibliography{helpers/MyRefLocal}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

>>>>

\usepackage[]{helpers/mypackages}

%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}

\bibliography{helpers/MyRefLocal}
\usepackage{helpers/myenvironments}
\usepackage{helpers/mycommands}

@@

VERSION 1

A central pivot in economics and decision theory is the definition of rationality as maximization of subjective preferences. The subjectivity of preferences plays a very important role. From an agent's standpoint, the amount of, say, money achieved does not necessarily coincide with the amount of subjective utility attained. The first to notice this difference between objective utility (goods, money, etc.) and subjective utility (preferences) was D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that subjective utility can be represented by an increasing concave function of objective utility. This can be considered the birth of modern decision theory.
Another important step was developed by Kahnemann and Tversky who introduced prospect theory in their famous paper from 1979. By means of experimental results they noticed that the decision maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent with respect to losses. Consequently, they conjectured that the subjective utility is better representable by an S-shaped utility function, concave in the positive domain of gains and convex in the negative domain of losses.
More recently, other possible definitions of the subjective utility function have been proposed and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin, etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas (e.g., Prisoner's dilemma, public goods game, etc.). By means of different subjective utility functions, it is then possible to describe/match the behavior observed in experiments and to explain it as rational (in the sense of maximizing the subjective utility). \\
With this approach it seems that every behavior, no matter how far from the maximization of objective utility, can be deemed as rational if we can only find a subjective utility that justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost) everything real is rational. The problem with this approach is that it seems that everything that is observed can be considered rational, given the adequate subjective utility. The notion of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rationality more tightly to the maximization of objective utility. In their classic works they argue that profit maximization is a reasonable assumption for characterizing outcomes in a competitive market because only firms behaving in a manner that is consistent with profit maximization will survive in the long run. If the previous accounts of rationality looked more descriptive, this definitely tastes more normative. In this sense, the normative notion of rationality is tightly linked to (and almost coincide with) the maximization of objective utility, which is supposed to drive the evolutionary process. It follows then that being normative in this sense seems to entail an identification of rationality with maximization of objective utility.

It seems that we are stuck between two extrema. On the one hand, if we want to have a descriptive notion of rationality, then it seems we can (almost) always define an appropriate subjective utility whose maximization matches the observed behavior. On the other hand, if we want to have a normative approach to rationality, then it seems we have to appeal to evolutionary arguments implying the understanding of rationality as maximization of objective utility.

The meta-game perspective that we introduce here can shed more light on this duality, and opens some space in between the two extrema. By keeping the evolutionary approach we want to retain a normative perspective on rationality, but our results show that (under certain conditions) evolution does not necessarily lead to identifying rationality with maximization of objective utility. More precisely, from the meta-game perspective many possible subjective transformations of the objective utility are eliminated by evolution, but we find some evolutionarily stable subjective utility maximizers whose utility does not coincide with the objective evolutionary utility. This result leaves some room in between the two extrema: we can understand rationality in a normative way, given by evolution, without having to discard all the possible subjective utility representations as irrational.

In particular, in the following we mainly deal with two types of agents: players whose utility function coincides with the objective utility, and players whose subjective utility is defined in terms of regret. The justification for this choice is both historical and behavioral. Among the possible competitors of utility maximization, regret has always been in a prominent position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes 
\begin{quote}
By explicitly incorporating regret, expected utility theory not only becomes a better descriptive predictor but also may become a more convincing guide for prescribing behavior to decision makers.
\end{quote}

\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match and explain the main experimental results of prospect theory, and claim that
\begin{quote}

[...] we shall challenge the idea that the conventional axioms constitute the only acceptable basis for rational choice under uncertainty. We shall argue that it is no less rational to act in accordance with regret theory, and that conventional expected utility theory therefore represents an unnecessarily restrictive notion of rationality. \cite{loosug82}

\end{quote}

...

END VERSION 1

VERSION 2

The importance of aiming at a richer evolutionary game theory has been recently emphasized by many works in theoretical biology and behavioral ecology [Fawcett\&al., McNamara, etc.]. The biological argument to embrace this goal is related to the criticisms towards the so-called behavioral gambit [Fawcett\&al.]. The behavioral gambit denotes the standard approach in evolutionary game theory that puts the focus of attention on the expressed behavior, and neglects the underlying mechanisms that generate that behavior. From this point of view, the benefit and the reliability coming from the use of evolutionary game theory for biological purposes is also under attack. There are good reasons to claim that focusing on the general psychological mechanisms instead of on behavior itself as the phenotype under evolutionary selection can be more faithful and insightful for natural biology and ecology [Fawcett\&al.].
Standard evolutionary game theory normally models the agents as playing a single fixed game, and as genetically predetermined to play fixed strategies, that consequently are the only target of natural selection in those frameworks. Standard models of evolutionary game theory cannot take charge of this shift of perspective until they will keep modelling a single game at a time, and until they will keep focusing on actions considered genetically encoded and fixed. ''Instead we should expect animals to have evolved a set of psychological mechanisms which enable them to perform well on average across a range of different circumstances''[Fawcett\&al.].

Attempts to overcome the behavioral gambit have recently been developed in economic literature too. In the last twenty years economists have been using evolutionary game-theoretical models to study evolution of preferences [cite], instead of evolution of behavior only. This approach is called indirect evolutionary approach, because players' behavior and actions stem from their subjective preferences, and it is aimed at investigating what subjective preferences are evolutionarily better off and robust with respect to natural selection. This shift from a direct evolutionary approach about observed actions and behavior to an indirect evolutionary approach takes only partly into account the behavioral gambit criticized by biologists and ecologists. The reason is that the models of evolution of preferences usually consider only one game at a time. This means that there is a single type of interaction that is supposed to drive the whole evolution of preferences: preferences are selected on the basis of one possible interaction only. Clearly, this approach does not take into consideration the variety of possible circumstances that animals are supposed to face during their lifetime.

The meta-game model we propose pays heed to both components of the behavioral gambit. We retain the indirect evolutionary approach, but our agents in the population play over a class of possible games $\mathcal{G}$. In this way we explicitly model a range of different circumstances, as wished by theoretical biologists.  Consequently, each agent in the population does not represent either a simple behavior or a subjective preference, but a more general player type that associates a subjective preference for each game in the class $\mathcal{G}$. The player type is supposed to encode the general psychological mechanisms that generate the behavior for any possible interaction in $\mathcal{G}$. These general psychological mechanisms then become the target of evolution, and they are selected based on the average fitness over a class of different interactions.

In particular, in the following we mainly deal with two types of agents: players whose subjective preference coincides with the evolutionary fitness, and players whose subjective preference is defined in terms of regret. It turns out that psychological mechanisms and subjective conceptualizations based on regret can outperform more veridical conceptualizations coinciding with the evolutionary fitness.
This is just an example of the kind of insights and results that we can gain by adopting a meta-game perspective and by shifting the focus of our attention to psychological mechanisms and subjective conceptualizations as the phenotype under selection.

-maybe to add: difference with learning. Fawcett\&al. talk more about differences in learning, whereas we model differences in subjective conceptualizations.

END VERSION 2
